Reactive SINDy: Discovering governing reactions from concentration data

### 2018-12-07 docs/manuscript_jcp.tex
comment out intertext (check if it breaks latexdiff)
break align in two separate ones, such that latexdiff works
wip

\begin{abstract}
The inner workings of a biological cell or a chemical reaction can
be rationalized by the network of reactions, whose structure reveals
the most important functional mechanisms. For complex systems, these
reaction networks are not known a priori and cannot be efficiently
computed with \emph{ab initio} methods, therefore an important approach
goal is to estimate effective reaction networks from observations,
such as time series of the main species. Reaction networks estimated
with standard machine learning techniques such as least-squares regression
may fit the observations, but will typically contain spurious reactions.
Here we extend the sparse identification of nonlinear dynamics (SINDy)
method to vector-valued ansatz functions, each describing a particular
reaction process. The resulting sparse tensor regression method ``reactive
SINDy'' is able to estimate a parsimonious reaction network. We illustrate
that a gene regulation network can be correctly estimated from observed
time series.
\end{abstract}

@@

We are observing the concentrations of $S$ chemical species in time
$t$: 
\begin{align}
\mathbf{x}(t)=\begin{pmatrix}x_{1}(t)\\
\vdots\\
x_{S}(t)
\end{pmatrix}\in\mathbb{R}^{S}.\label{eq:concentration_vector}
\end{align}
We assume that their dynamics are governed by classical reaction-rate
equations subject to the law of mass action. A general expression
for the change of concentration of reactant $s$ as a result of order-0
reactions (creation or annihilation), order-1 reactions (transitions
of other species into $s$ or transitions of $s$ into other species),
order-2 reactions (production or consumption of $s$ by the encounter
of two species), etc, is given by:

@@

Here we model the MAPK pathway with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network reads
\begin{align*}
\intertext{Activations:}
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    }
\intertext{Deactivations:}
\mathrm{MAPKKK*}    \mathrm{      }  & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}     \mathrm{      }  & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}      \mathrm{      }  & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}        \mathrm{      }  & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the response, i.e. the concentration $[\mathrm{TF*}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.

>>>>

\begin{abstract}
The inner workings of a biological cell or a chemical reactor can
be rationalized by the network of reactions, whose structure reveals
the most important functional mechanisms. For complex systems, these
reaction networks are not known a priori and cannot be efficiently
computed with \emph{ab initio} methods, therefore an important approach
goal is to estimate effective reaction networks from observations,
such as time series of the main species. Reaction networks estimated
with standard machine learning techniques such as least-squares regression
may fit the observations, but will typically contain spurious reactions.
Here we extend the sparse identification of nonlinear dynamics (SINDy)
method to vector-valued ansatz functions, each describing a particular
reaction process. The resulting sparse tensor regression method ``reactive
SINDy'' is able to estimate a parsimonious reaction network. We illustrate
that a gene regulation network can be correctly estimated from observed
time series.
\end{abstract}

@@

We are observing the concentrations of $S$ chemical species in time
$t$: 
\begin{align}
\mathbf{x}(t)=\begin{pmatrix}x_{1}(t)\\
\vdots\\
x_{S}(t)
\end{pmatrix}\in\mathbb{R}^{S}.\label{eq:concentration_vector}
\end{align}
We assume that their dynamics are governed by classical reaction-rate
equations subject to the law of mass action. A general expression
for the change of concentration of reactant $s$ as a result of order-0
reactions (creation), order-1 reactions (transitions
of other species into $s$, transitions of $s$ into other species, or annihilation),
order-2 reactions (production or consumption of $s$ by the encounter
of two species), etc, is given by:

@@

Here we model the MAPK pathway with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network consists of activation/phosphorylation reactions
\begin{align*}
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    } \\
\end{align*}
and deactivation/dephosphorylation reactions
\begin{align*}
\mathrm{MAPKKK*}                     & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}                      & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}                       & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}                         & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the response, i.e. the concentration $[\mathrm{TF*}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.
